Concurrent Programming in Harmony

### 2021-11-01 book/paper.tex
Working on book

\item \autoref{ch:synch} presents Harmony modules
that support locks and other synchronization primitives.

@@

Finally, methods can have local variables.
Different from Python, Harmony local variables must be
explicitly bound within a scope.
You have already seen examples: the arguments to the method,
the variable \textit{result}, and the bound variable (or
variables) within a \texttt{for} statement.
The \texttt{let} statement allows declaring additional
local variables.  For example:
\texttt{let} $x = 3$: $y$ $+$$=$ $x$ adds 3 to the global variable
$y$.
See \autoref{sec:vars} for more information.

@@

The code in \autoref{fig:lockspec} is taken from Harmony's \texttt{synch}
module.  \autoref{fig:cssynch} shows how locks may be used.  Note that
the code of \autoref{fig:lockspec} is executable in Harmony.
However, the \texttt{atomically} keyword is not available in general
programming languages and should not be used for implementation.
Peterson's algorithm is an implementation of a lock, although only for
two processes.  In the following chapters we will look at more general
ways of implementing locks using atomic constructions that are usually
available in the underlying hardware.

In Harmony, any statement can be preceded by the \texttt{atomically}
keyword.  It means that statement as a whole is to be executed atomically.
The \texttt{atomically} keyword can be used to specify the behavior of
methods such as \texttt{acquire} and \texttt{release}.  But note that
an actual executable program such as in \autoref{fig:cssynch} should not
use the keyword.  In fact, if we want to make the entire program
executable, we have to also show how \texttt{acquire} and \texttt{release}
are implemented, not just how they are specified.
\autoref{ch:spinlock} presents such an implementation.

The code in \autoref{fig:lockspec} also uses the Harmony \texttt{when}
statement.  A \texttt{when} statement waits until a time in which
condition holds (not necessarily the first time) and then executes the
statement block.  The \texttt{await} \textit{condition} statement is
the same as \texttt{when} \textit{condition}: \texttt{pass}.
Combined with the \texttt{atomically} keyword, the entire statement
is executed atomically at a time that the condition holds.

@@

To support implementing locks,
many CPUs have an atomic ``test-and-set'' (TAS)
\index{test-and-set}%
\index{TAS}%
operation.
Method \texttt{tas} in \autoref{fig:tas} shows its specification.
Here $s$ points to a shared Boolean variable and $p$
to a private Boolean variable, belonging to some thread.
The operation copies the value of the shared variable to the
private variable (the ``test'')
and then sets the shared variable to \textbf{True} (``set'').

@@

An important invariant, $\mathcal{I}_1$, of the program is that at any time at most
one of these variables is \textbf{False}.
Another invariant, $\mathcal{I}_2(i)$, is that if thread $i$ is executing between
lines $18 \cdots 22$ (which includes the critical section),
then \textit{private}[$i$] == \textbf{False}.
Between the two (i.e., $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$),
it is clear that only one thread can be in the
critical section at the same time.
For those who read through the previous chapter:
$\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$ is an inductive invariant.

To see that invariant $\mathcal{I}_1$ is maintained, note that
\texttt{!}$p$ == \textbf{True} upon entry of \texttt{tas}
(because of the condition on the \textbf{while} loop that the
\texttt{tas} method is invoked in).
There are two cases:
\begin{enumerate}
\item \texttt{!}$s$ is \textbf{False} upon entry to \texttt{tas}.
Then upon exit \texttt{!}$p$ == \textbf{False} and \texttt{!}$s$ == \textbf{True},
maintaining the invariant.
\item \texttt{!}$s$ is \textbf{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section
because we can assume, by the induction hypothesis, that $\textit{private}[i]$ is
\textbf{True} just before exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) thread $i$ only proceeds to the critical
section if \textit{private}[$i$] is \textbf{False}, and (ii) no other thread modifies
\textit{private}[$i$].

@@

\chapter{Blocking}
\label{ch:synch}
\index{lock}%
\index{synch module}%

\begin{figure}
\begin{code}
\harmonysource{taslock}
\end{code}
\caption{\harmonylink{code/taslock.hny} Implementation of a spinlock
using test-and-set.}
\label{fig:tas2}
\end{figure}

In \autoref{fig:tas} we have shown a solution based on a shared
variable and a private variable for each thread.   The private
variables themselves are actually implemented as shared variables,
but they are accessed only by their respective threads.
A thread usually does not keep explicit track of whether it has a lock
or not, because it is implied by the control flow of the program---a
thread implicitly \emph{knows} that when it is executing in a critical
section it has the lock.
There is no need to keep \textit{private} as a shared
variable---we only did so to be able to show and check the invariants.
\autoref{fig:tas2} shows a more straightforward implementation of spinlock.
The solution is similar to the na\"{i}ve solution of \autoref{fig:naivelock},
but uses test-and-set to check and set the lock variable atomically.
The lock is also cleared in an atomic statement to prevent a data race.
This approach is general for any number of threads.

In Harmony, atomic statements allow you to \emph{implement} your own
atomic primitives such as test-and-set.  Other common examples
include compare-and swap and fetch-and-add.  Atomic statements
are not intended to replace locks or other synchonization primitives.
When solving synchronization problems you should not directly use
atomic statements but use the synchronization primitives that are available
to you.  But if you want to design a new synchronization primitive, then
use \textbf{atomically} by all means.
You can also use atomic statements in your test code.
In fact, as mentioned before, \textbf{assert} statements are executed atomically.

\begin{figure}
\begin{code}
\harmonysource{lockintf}
\end{code}
\caption{\harmonylink{modules/synch.hny} The binary semaphore interface and implementation in the \texttt{synch} module.}
\label{fig:spinlocks}
\end{figure}

@@

@@

\index{binary semaphore}
Locks are a special case of \emph{binary semaphores}.
Like locks, binary semaphores are either in an \emph{acquired} state
or a \emph{released} state.
Threads trying to acquire an already acquired binary semaphore
have to wait until the binary semaphore is released.
It is illegal to release a binary semaphore that is in a released
state.
A lock is a binary semaphore that, by convention, is initialized to a
released state, and is always released by the same thread that acquired it.
A binary semaphore can be initialized to an acquired state and can be
released by threads other than the thread that last acquired the
binary semaphore.

Harmony has a module called \texttt{synch} that includes support for
binary semaphores and locks.
\autoref{fig:spinlocks} shows how they are implemented, and
\autoref{fig:incfixed} gives an example of how they may be used,
in this case to fix the program of \autoref{fig:inc}.
Notice that the module hides the implementation of
binary semaphores and locks.
The \texttt{synch} module includes a variety of other useful
synchronization primitives, which will be discussed in later
chapters.
Also notice the use of the sequentially consistent \textit{done}
variable for testing purposes.

@@

We call a thread \emph{blocked}
\index{blocked thread}%
if a thread cannot change the state or terminate unless
another thread changes the state first.
A thread trying to
acquire a test-and-set spinlock held by another thread is a good example
of a thread being blocked.
The only way forward is if the other thread releases the spinlock.
A thread that is in an infinite loop is also considered blocked.
% although Harmony can often detect infinite loops specifically.

\begin{figure}
\begin{code}
\harmonysource{locksusp}
\end{code}
\caption{\harmonylink{modules/synchS.hny} The binary semaphore interface in the \texttt{synchS} module uses suspension.}
\label{fig:suspension}
\end{figure}

@@

There is a second version of the \texttt{synch} module, called \texttt{synchS},
that uses suspension instead of spinlocks.
\autoref{fig:suspension} shows the same interface
as \autoref{fig:spinlocks} implemented using suspension.
The interface is implemented as follows:
\begin{itemize}
\item A binary semaphore maintains both a boolean indicating whether the
binary semaphore is currently acquired and a list of contexts of threads that want to
acquire the binary semaphore.
\item
\texttt{acquire}()
\index{acquire}%
acquires the binary semaphore if available and suspends the invoking thread if not.
In the latter case, the context of the thread is added to the end of the list of contexts.
Note that \textbf{stop} is called within an atomic statement block---this is
the only exception to such an atomic statement block running to completion.
While the
thread is running no other threads can run, but when the thread suspends itself
other threads can run.
\item
\texttt{release}()
\index{release}%
checks to see if any threads are waiting to acquire the binary semaphore.
If so, it uses the \texttt{head} and \texttt{tail}
methods from the \texttt{list} module (see \autoref{ap:module})
to resume the first thread that got
suspended and to remove its context from the list.
\end{itemize}
Selecting the first thread is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
\texttt{choose} to pick an arbitrary one.  Selecting the first is a common
choice in binary semaphore implementations as it prevents \emph{starvation}:
\index{starvation}%
every thread
gets a chance to acquire the binary semaphore (assuming every thread eventually releases
it).  \autoref{ch:starvation} will talk more about starvation and how
to prevent it.

Harmony allows you to select which version of the \texttt{synch} module you would
like to use with the \texttt{-m} flag.
\index{module}%
For example,

\begin{code}
\begin{verbatim}
harmony -m synch=synchS x.hny
\end{verbatim}
\end{code}

runs the file \texttt{x.hny} using the suspension version of the \texttt{synch} module.
You will find that using the \texttt{synchS} module often leads to
the model checker searching a
significantly larger state space than using the \texttt{synch} module.
Part of the reason is that the \texttt{synchS} module keeps track of the order
in which threads wait to acquire a binary semaphore, while the \texttt{synch} module does not.

>>>>

\item \autoref{ch:synch} looks at various ways in which the lock specification
in \autoref{ch:specification} can be implemented.

@@

Finally, methods can have local variables.
Different from Python, Harmony local variables must be
explicitly bound within a scope.
Method variables are either mutable or immutable.
The arguments to a method and the bound variable (or
variables) within a \texttt{for} statement are immutable;
the variable \textit{result} is mutable.
Using the \texttt{var} statement, new mutable local variables can be
declared.
For example, \texttt{var} $x = 3$ declares a new mutable local variable
$x$.
The \texttt{let} statement allows declaring new immutable local variables.
For example:
\texttt{let} $x = 3$: $y$ $+$$=$ $x$ adds 3 to the global variable
$y$.
See \autoref{sec:vars} for more information.

@@

The code in \autoref{fig:lockspec} is similar to the code in
Harmony's \texttt{synch} module.
(The module generalizes locks to \emph{binary semaphores}, but the
lock interface is the same.)
\autoref{fig:cssynch} shows how locks may be used to implement
a critical section.
\autoref{fig:incfixed} gives an example of how locks may be used
to fix the program of \autoref{fig:inc}.

Note that the code of \autoref{fig:lockspec} is executable in Harmony.
However, the \texttt{atomically} keyword is not available in general
programming languages and should not be used for implementation.
Peterson's algorithm is an implementation of a lock, although only for
two processes.  In the following chapters, we will look at more general
ways of implementing locks using atomic constructions that are usually
available in the underlying hardware.

In Harmony, any statement can be preceded by the \texttt{atomically}
keyword.  It means that statement as a whole is to be executed atomically.
The \texttt{atomically} keyword can be used to specify the behavior of
methods such as \texttt{acquire} and \texttt{release}.  But note that
an actual executable program such as in \autoref{fig:cssynch} should not
use the \texttt{atomically} keyword.
In fact, if we want to make the entire program executable on hardware,
we have to also show how \texttt{Lock}, \texttt{acquire}, and \texttt{release}
are implemented, not just how they are specified.
\autoref{ch:spinlock} presents such an implementation.

The code in \autoref{fig:lockspec} also uses the Harmony \texttt{when}
statement.  A \texttt{when} statement waits until a time in which
condition holds (not necessarily the first time) and then executes the
statement block.  The ``\texttt{await} \textit{condition}'' statement is
the same as ``\texttt{when} \textit{condition}: \texttt{pass}''.
Combined with the \texttt{atomically} keyword, the entire statement
is executed atomically at a time that the condition holds.

% In Harmony, atomic statements allow you to \emph{implement} your own
% atomic primitives such as test-and-set.  Other common examples
% include compare-and swap and fetch-and-add.
Atomic statements
are not intended to replace locks or other synchonization primitives.
When solving synchronization problems you should not directly use
atomic statements but use the synchronization primitives that are available
to you.  But if you want to \emph{specify} a new synchronization primitive,
then use \textbf{atomically} by all means.
You can also use atomic statements in your test code.
In fact, as mentioned before, \textbf{assert} statements are included to
test if certain conditions hold in every execution and are executed
atomically.

@@

To support implementing locks,
many CPUs have an atomic ``test-and-set'' (TAS)
\index{test-and-set}%
\index{TAS}%
operation.
Method \texttt{test\_and\_set} in \autoref{fig:tas} shows its specification.
Here $s$ points to a shared Boolean variable and $p$
to a private Boolean variable, belonging to some thread.
The operation copies the value of the shared variable to the
private variable (the ``test'')
and then sets the shared variable to \textbf{True} (``set'').

@@

An important invariant, $\mathcal{I}_1$, of the program is that at any time at most
one of these variables is \textbf{False}.
Another invariant, $\mathcal{I}_2(i)$, is that if thread $i$ is executing between
lines $18 \cdots 22$ (which includes the critical section),
then \textit{private}[$i$] == \textbf{False}.
Between the two (i.e., $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$),
it is clear that only one thread can be in the
critical section at the same time.
For those who read through \autoref{ch:peterson}
$\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$ is an inductive invariant.

To see that invariant $\mathcal{I}_1$ is maintained, note that
\texttt{!}$p$ == \textbf{True} upon entry of \texttt{test\_and\_set}
(because of the condition on the \textbf{while} loop that the
\texttt{test\_and\_set} method is invoked in).
There are two cases:
\begin{enumerate}
\item \texttt{!}$s$ is \textbf{False} upon entry to \texttt{test\_and\_set}.
Then upon exit \texttt{!}$p$ == \textbf{False} and \texttt{!}$s$ == \textbf{True},
maintaining the invariant.
\item \texttt{!}$s$ is \textbf{True} upon entry to \texttt{test\_and\_set}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section
because we can assume, by the induction hypothesis, that $\textit{private}[i]$ is
\textbf{True} just before exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) thread $i$ only proceeds to the critical
section if \textit{private}[$i$] is \textbf{False}, and (ii) no other thread modifies
\textit{private}[$i$].

@@

\chapter{Lock Implementations}
\label{ch:synch}
\index{lock}%
\index{synch module}%

\begin{figure}
\begin{code}
\harmonysource{taslock}
\end{code}
\caption{\harmonylink{code/taslock.hny} Implementation of the lock
specification in \autoref{fig:lockspec} using a spinlock based on
test-and-set.}
\label{fig:tas2}
\end{figure}

\begin{figure}
\begin{code}
\harmonysource{ticket}
\end{code}
\caption{\harmonylink{code/ticket.hny} Implementation of the lock
specification in \autoref{fig:lockspec} using a ticket lock.}
\label{fig:ticket}
\end{figure}

@@

In \autoref{fig:tas} we have shown a lock implementation based on a shared
variable and a private variable for each thread.   The private
variables themselves are actually implemented as shared variables,
but they are accessed only by their respective threads.
A thread usually does not keep explicit track of whether it has a lock
or not, because it is implied by the control flow of the program---a
thread implicitly \emph{knows} that when it is executing in a critical
section it has the lock.
There is no need to keep \textit{private} as a shared
variable---we only did so to be able to show and check the invariants.
\autoref{fig:tas2} shows a more straightforward implementation of a spinlock.
% The solution is similar to the na\"{i}ve solution of \autoref{fig:naivelock},
% but uses test-and-set to check and set the lock variable atomically.
The lock is also cleared in an atomic statement to prevent a data race.
This approach is general for any number of threads.

You can test the spinlock with the program in \autoref{fig:cssynch}
using the command \texttt{harmony -m synch=taslock code/cssynch.hny}.
The \texttt{-m} flag tells Harmony to use the \texttt{taslock.hny} file
rather than the standard \texttt{synch} module (which only contains
a specification of the lock methods).

\index{starvation}%
The spinlock implementation suffers potentially from \emph{starvation}:
an unlucky thread may never be able to get the lock while other threads
successfully acquire the lock one after another.  It could even happen
with just two threads: one thread might successfully acquire the lock
repeatedly in a loop, while another thread is never lucky enough the
acquire the thread in between.
\autoref{ch:starvation} will talk more about starvation and how
to prevent it.

A \emph{ticket lock} (\autoref{fig:ticket} is an implementation of a
lock that prevents starvation using an atomic \emph{fetch-and-increment}
operator.  It is inspired by European bakeries.  A European bakery has a
clearly displayed counter (usually just two digits) and a ticket dispenser.
Tickets are numbered 0 through 99 and repeat over and over again
(in the case of a two digit counter).
When a customer walks into the bakery, they draw a number from the dispenser
and wait until their number comes up.
Every time a customer has been helped, the counter is incremented.
(Note that this only works if there can be no more than 100 customers in
the bakery at a time.)

\autoref{fig:ticket} similarly uses two variables for a lock,
\textit{counter} and \textit{dispenser}.  When a thread acquires the
lock, it fetches the current dispenser value and increments it
modulo \texttt{MAX\_THREADS}, all in one atomic operation.
In practice, \texttt{MAX\_THREADS} would
be a number like $2^32$ or $2^64$, but since the Harmony model checker
checks every possible state, limiting it to a small number such as done
here will significantly reduce the time to model check a Harmony program.
Plus it is easier to check that it fails when you run it with more than
\texttt{MAX\_THREADS} threads.
You can test the implementation using the command
\texttt{harmony -m synch=ticket code/cssynch.hny}.
To see it fail, try
\texttt{harmony -c NTHREADS=10 -m synch=ticket code/cssynch.hny}.

@@

\begin{comment}
\index{binary semaphore}
Locks are a special case of \emph{binary semaphores}.
Like locks, binary semaphores are either in an \emph{acquired} state
or a \emph{released} state.
Threads trying to acquire an already acquired binary semaphore
have to wait until the binary semaphore is released.
It is illegal to release a binary semaphore that is in a released
state.
A lock is a binary semaphore that, by convention, is initialized to a
released state, and is always released by the same thread that acquired it.
A binary semaphore can be initialized to an acquired state and can be
released by threads other than the thread that last acquired the
binary semaphore.
\end{comment}

@@

We now turn to a radically different way of implementing locks, one
that is commonly provided by operating systems to user processes.
We call a thread \emph{blocked}
\index{blocked thread}%
if a thread cannot change the state or terminate unless
another thread changes the state first.
A thread trying to acquire a lock held by another thread is a good example
of a thread being blocked.
The only way forward is if the other thread releases the lock.
A thread that is in an infinite loop is also considered blocked.

\begin{figure}
\begin{code}
\harmonysource{locksusp}
\end{code}
\caption{\harmonylink{modules/synchS.hny} The lock implementation in the \texttt{synchS} module uses suspension.}
\label{fig:suspension}
\end{figure}

@@

\autoref{fig:suspension} shows the lock interface using suspension.
It is implemented as follows:
\begin{itemize}
\item A lock maintains both a boolean indicating whether the
lock is currently acquired and a list of contexts of threads that want to
acquire the lock.
\item
\texttt{acquire}()
\index{acquire}%
acquires the lock if available and suspends the invoking thread if not.
In the latter case, the context of the thread is added to the end of the list of contexts.
Note that \textbf{stop} is called within an atomic statement block---this is
the only exception to such an atomic statement block running to completion.
While the
thread is running no other threads can run, but when the thread suspends itself
other threads can run.
\item
\texttt{release}()
\index{release}%
checks to see if any threads are waiting to acquire the lock.
If so, it uses the \texttt{head} and \texttt{tail}
methods from the \texttt{list} module (see \autoref{ap:module})
to resume the first thread that got
suspended and to remove its context from the list.
\end{itemize}
Selecting the first thread is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
\texttt{choose} to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents starvation.

You will find that using the \emph{implementation} of a lock instead
of the \emph{specification} of a lock (in the \texttt{synch} module)
often leads to the model checker searching a
significantly larger state space.
Thus it makes sense to model check larger programs in a modular fashion:
model check one module implementation at a time, using
specifications for the other modules.
